# Data Processing

Load the standard libraries we use to do work

```{r, warning=FALSE, message=FALSE}
# bread-and-butter
library(tidyverse) # the tidyverse
library(viridis) # viridis colors
library(harrypotter) # hp colors
library(RColorBrewer) # brewer colors
library(scales) # work with number and plot scales
library(latex2exp)

# visualization
library(mapview) # interactive html maps
library(kableExtra) # tables
library(patchwork) # combine plots

# spatial analysis
library(terra) # raster
library(sf) # simple features
library(lidR) # lidar data
library(rgl) # 3d plots
library(cloud2trees) # the cloud2trees
```

```{r pkg-ld, include=F, warning=F, message=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results = 'hide'
  , fig.width = 10.5
  , fig.height = 7
)
# option to put satellite imagery as base layer of mapview maps
  mapview::mapviewOptions(
    homebutton = FALSE
    # , basemaps = c("Esri.WorldImagery","OpenStreetMap")
    , basemaps = c("OpenStreetMap", "Esri.WorldImagery")
  )
# clean session
remove(list = ls())
gc()
```

## Slash Pile Vector Data

```{r}
slash_piles_sf <- sf::st_read("D:\\PFDP_Data\\PFDP_SlashPiles\\SlashPiles.shp") %>% 
  dplyr::rename_with(tolower)
```

what? 

```{r}
slash_piles_sf %>% dplyr::glimpse()
```

where?

```{r}
mapview::mapview(slash_piles_sf, zcol = "comment", layer.name = "slash piles")
```

## Point Cloud Data

Let's check out the point cloud data we got using UAS-SfM methods

```{r}
# directory with the downloaded .las|.laz files
f <- "D:\\PFDP_Data\\p4pro_images\\P4Pro_06_17_2021_half_half_optimal\\2_densification\\point_cloud"
# is there data?
list.files(f, pattern = ".*\\.(laz|las)$") %>% length()
# what files are in here?
list.files(f, pattern = ".*\\.(laz|las)$")[1]
```

what information does `lidR` read from the catalog?

```{r}
ctg_temp <- lidR::readLAScatalog(f)
# set the processing options
lidR::opt_progress(ctg_temp) <- F
lidR::opt_filter(ctg_temp) <- "-drop_duplicates"
lidR::opt_select(ctg_temp) <- "xyziRGB"
# huh?
ctg_temp
```

that's a lot of points...can an ordinary laptop handle it? we'll find out.

We'll plot our point cloud data tiles real quick to orient ourselves

```{r}
ctg_temp %>% 
  purrr::pluck("data") %>% 
  mapview::mapview(popup = F, layer.name = "point cloud tile")
```

## Check out one pile

```{r}
las_temp <- lidR::clip_roi(
  ctg_temp
  # biggest mechanical
  , slash_piles_sf %>%
    dplyr::filter(tolower(comment)=="mechanical pile") %>% 
    dplyr::arrange(desc(diameter)) %>% 
    dplyr::slice(1) %>% 
    sf::st_zm() %>% 
    sf::st_buffer(10, endCapStyle = "SQUARE") %>% 
    sf::st_transform(lidR::st_crs(ctg_temp))
)
```

what did we get?

```{r}
las_temp@data %>% dplyr::glimpse()
```

plot a sample

```{r,echo=FALSE,message=FALSE,warning=FALSE}
## if want to get current rgl parameters
# rgl::par3d()$zoom
# rgl::par3d()$FOV
# rgl::par3d()$userMatrix %>% c()
# rgl::par3d()$windowRect
## set up for printing rgl graphics
r3dDefaults <- rgl::r3dDefaults
m  <- structure(c(
  -0.39951634,0.09665099, -0.91161686,0.00000000, -0.90745372,0.09938181,0.40822852,0.00000000,0.13005376,0.99034417,0.04800174,0.00000000,0.00000000,0.00000000
,0.00000000,1.00000000
), .Dim = c(4L, 4L))
r3dDefaults$FOV <- 30
r3dDefaults$userMatrix <- m
r3dDefaults$zoom <- 0.78
r3dDefaults$windowRect <- c(0,23,1536,864)
rgl::setupKnitr(autoprint = TRUE)
```

```{r, rgl = TRUE}
las_temp %>% 
  lidR::plot(
    color = "Z", bg = "white", legend = F
   , pal = harrypotter::hp(n=50, house = "gryffindor")
  )
```

make a gif

```{r}
library(magick)
if(!file.exists(file.path("../data/", "pile_z.gif"))){
  rgl::close3d()
  lidR::plot(
    las_temp, color = "Z", bg = "white", legend = F
   , pal = harrypotter::hp(n=50, house = "gryffindor")
  )
  rgl::movie3d( rgl::spin3d(), duration = 10, fps = 10 , movie = "pile_z", dir = "../data/")
  rgl::close3d()
}
```

```{r}
library(magick)
if(!file.exists(file.path("../data/", "pile_rgb.gif"))){
  rgl::close3d()
  lidR::plot(
    las_temp, color = "RGB", bg = "white", legend = F
  )
  rgl::movie3d( rgl::spin3d(), duration = 10, fps = 10 , movie = "pile_rgb", dir = "../data/")
  rgl::close3d()
}
```

## Load RGB orthomosaic rasters

Orthomosaic `tif` files from the UAS flight imagery that were created in [Agisoft Metashape](https://www.agisoft.com/) are loaded and stitched together via `terra::mosaic`.

```{r ortho-ld, results = 'hide'}
  # read list of orthos
  ortho_list_temp <- list.files(
    "D:\\PFDP_Data\\p4pro_images\\P4Pro_06_17_2021_half_half_optimal\\3_dsm_ortho\\2_mosaic"
    , pattern = "*\\.(tif|tiff)$", full.names = T)[] %>% 
    purrr::map(function(x){terra::rast(x)})
  
  # ortho_list_temp[[1]] %>%
  #   terra::aggregate(20) %>%
  #   terra::plotRGB(r = 1, g = 2, b = 3, stretch = "hist", colNA = "transparent")
  
  ####### ensure the resolution of the rasters matches 
    # terra::res(ortho_list_temp[[1]])
    
    ## function
    change_res_fn <- function(r, my_res=1){
      r2 <- r
      terra::res(r2) <- my_res
      r2 <- terra::resample(r, r2)
      return(r2)
    }
    ## apply the function
    ortho_list_temp <- 1:length(ortho_list_temp) %>% 
      purrr::map(function(x){change_res_fn(ortho_list_temp[[x]], my_res=0.25)})
    
    # terra::res(ortho_list_temp[[1]])
    
    # ortho_list_temp[[1]] %>%
    #   terra::aggregate(2) %>%
    #   terra::plotRGB(r = 1, g = 2, b = 3, stretch = "hist", colNA = "transparent")

  ######## mosaic the raster list 
    ortho_rast <- terra::mosaic(
      terra::sprc(ortho_list_temp)
      , fun = "min" # min only thing that works
    ) 
    
  # ortho_rast %>%
  #   terra::aggregate(4) %>%
  #   terra::plotRGB(r = 1, g = 2, b = 3, stretch = "lin", colNA = "transparent")
```

ortho plot fn

```{r}
######################################################################################
# function to plot ortho + stand
######################################################################################
ortho_plt_fn = function(stand = las_ctg_dta %>% sf::st_union() %>% sf::st_as_sf(), buffer = 20){
# convert to stars
  ortho_st <- ortho_rast %>%  
    terra::subset(subset = c(1,2,3)) %>%
    terra::crop(
      stand %>% sf::st_buffer(buffer) %>% terra::vect()
    ) %>% 
    terra::aggregate(fact = 2, fun = "mean", na.rm = T) %>% 
    stars::st_as_stars()
  
  # convert to rgb
  ortho_rgb <- stars::st_rgb(
    ortho_st[,,,1:3]
    , dimension = 3
    , use_alpha = FALSE
    # , stretch = "histogram"
    , probs = c(0.005, 0.995)
    , stretch = "percent"
  )
  # ggplot
  plt_rgb <- ggplot() +
    stars::geom_stars(data = ortho_rgb[]) +
    scale_fill_identity(na.value = "transparent") + # !!! don't take this out or RGB plot will kill your computer
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(
      x = ""
      , y = ""
    ) +
    theme_void()
  
  # return(plt_rgb)
  # combine all plot elements
  plt_combine = plt_rgb +
    # geom_sf(
    #   data = stand
    #   , alpha = 0
    #   , lwd = 1.5
    #   , color = "gray22"
    # ) +
    theme(
      legend.position = "top" # c(0.5,1)
      , legend.direction = "horizontal"
      , legend.margin = margin(0,0,0,0)
      , legend.text = element_text(size = 8)
      , legend.title = element_text(size = 8)
      , legend.key = element_rect(fill = "white")
      # , plot.title = ggtext::element_markdown(size = 10, hjust = 0.5)
      , plot.title = element_text(size = 8, hjust = 0.5, face = "bold")
      , plot.subtitle = element_text(size = 6, hjust = 0.5, face = "italic")
    )
  return(plt_combine)
}
```

plot an example slash pile RGB image

```{r}
stand_temp = slash_piles_sf %>%
    dplyr::filter(tolower(comment)=="mechanical pile") %>% 
    dplyr::arrange(desc(diameter)) %>% 
    dplyr::slice(1) %>% 
    sf::st_zm() %>% 
    sf::st_buffer(10, endCapStyle = "SQUARE") %>% 
    sf::st_transform(terra::crs(ortho_rast))
# check it with the ortho
ortho_plt_fn(stand = stand_temp)
ggsave("../data/pile_rgb.jpeg", height = 5, width = 5)
```

### Example ratio-based index

let’s define a general function for a ratio based (vegetation) index

```{r}
spectral_index_fn <- function(rast, layer1, layer2) {
  bk <- rast[[layer1]]
  bi <- rast[[layer2]]
  vi <- (bk - bi) / (bk + bi)
  return(vi)
}
```

The Green-Red Vegetation Index (GRVI) uses the reflectance of green and red bands to assess vegetation health and identify ground cover types. The formula is GRVI = (green - red) / (green + red). Higher GRVI values indicate healthy vegetation, while negative values suggest soils, and values near zero may indicate water or snow.

```{r}
grvi_rast <- spectral_index_fn(rast = ortho_rast, layer1 = 2, layer2 = 1)
terra::plot(grvi_rast)
```

let's check the GRVI for a pile

```{r}
# check it with the ortho
grvi_rast %>% 
  terra::crop(stand_temp %>% sf::st_buffer(20)) %>% 
  terra::as.data.frame(xy=T) %>% 
  dplyr::rename(f=3) %>% 
  ggplot2::ggplot() +
  ggplot2::geom_tile(ggplot2::aes(x=x,y=y,fill = f), color = NA) +
  ggplot2::labs(fill="CHM (m)") +
  # harrypotter::scale_fill_hp("slytherin") +
  scale_fill_gradient2(low = "black", high = "forestgreen") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  ggplot2::theme_void() +
  theme(
      legend.position = "none" # c(0.5,1)
      , legend.direction = "horizontal"
      , legend.margin = margin(0,0,0,0)
      , legend.text = element_text(size = 8)
      , legend.title = element_text(size = 8)
      , legend.key = element_rect(fill = "white")
      # , plot.title = ggtext::element_markdown(size = 10, hjust = 0.5)
      , plot.title = element_text(size = 10, hjust = 0.5, face = "bold")
      , plot.subtitle = element_text(size = 8, hjust = 0.5, face = "italic")
    )

ggsave("../data/pile_grvi.jpeg", height = 5, width = 5)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
remove(f, r3dDefaults, m)
gc()
```

## Study area imagery

let's look at the RGB imagery and pile locations

```{r}
# get the base plot
plt_rgb_ortho <- ortho_plt_fn(
  slash_piles_sf %>% 
  sf::st_bbox() %>% 
  sf::st_as_sfc() %>% 
  sf::st_buffer(50) %>% 
  sf::st_transform(terra::crs(ortho_rast))
)
# add pile locations
plt_rgb_ortho +
  ggplot2::geom_sf(
    data = slash_piles_sf %>% sf::st_transform(terra::crs(ortho_rast))
    , ggplot2::aes() # size = diameter
    , shape = 1
    , color = "firebrick"
  ) +
  ggplot2::theme(legend.position = "none")
```

notice these are point measurements of plot locations and the points are not precisely in the center of the pile. notice also there are piles in the imagery that were not measured (e.g. upper-left corner)

let's make a panel of plots for each pile

```{r}
p_fn_temp <- function(rn, df = slash_piles_sf, buff = 6.5, crs = terra::crs(ortho_rast)) {
  # scale the buffer based on the largest
  sc <- buff/max(df$diameter, na.rm = T)
  d <- df %>%
    dplyr::arrange(tolower(comment), desc(diameter)) %>% 
    dplyr::slice(rn) %>% 
    sf::st_zm() 
  d <- d  %>% 
    sf::st_buffer(
      dplyr::case_when(
        is.null(d$diameter) | is.na(d$diameter) ~ buff
        , d$diameter <= 0 ~ buff
        , T ~ sc*d$diameter
      )
      , endCapStyle = "ROUND"
    ) %>% 
    sf::st_transform(crs)
  # plt
  ortho_plt_fn(d) + 
    ggplot2::geom_sf(data = d, fill = NA, color = "firebrick") +
    ggplot2::labs(
      subtitle = paste0(
        tolower(d$comment)
        , "\ndiam. = "
        , scales::comma(d$diameter, accuracy = 0.1)
        # , ", ht. = "
        # , scales::comma(d$height, accuracy = 0.1)
      )
    )
}
# add pile locations
plt_list_rgb <- 1:nrow(slash_piles_sf) %>% 
  purrr::map(p_fn_temp)
```

plot tiles 

```{r, height = 10.5, width = 8}
patchwork::wrap_plots(
  sample(
    plt_list_rgb, size = as.integer(nrow(slash_piles_sf)/3))
  , ncol = 5
)
ggsave("../data/pile_tiles_rgb.jpeg", height = 10.5, width = 8)
```

we have point locations of piles. object-based image analysis will require polygon data so that we have sets of pixels (i.e. "objects") to work with for training our model.

notice that defining the pile "objects" with this point data is going to be challenging given the point locations appear to not have been measured at the center of each pile. if they had, we could use the point location and then buffer by the pile diameter to get the pile extent.

another challenge will be training a model with the different lighting conditions in the imagery (e.g. piles in shadows or under tree crowns)

