# RGB Classification

We'll use the RGB imagery alone to explore unsupervised and supervised classification algorithms.

Ultimately, it is likely we will use an object-based image analysis (OBIA) approach. OBIA is a technique (or a set of techniques) used to analyze digital images that was developed relatively recently in comparison to ‘classic’ pixel-based image approaches ([Burnett and Blaschke 2003](https://doi.org/10.1016/S0304-3800(03)00139-X)). The OBIA approach for ecological applications is described by [Goncalves et al. (2019)](https://doi.org/10.1016/j.jag.2018.11.011) who also introduce the `SegOptim` [package](https://github.com/joaofgoncalves/SegOptim)

In future sections we will explore, integration of spectral data with the 3D point cloud data (i.e. "data fusion") to test distinguishing the woody material in slash piles from surrounding vegetation or soil. Field-measured slash piles will be used as the ground truth data to perform a confusion matrix-based validation accuracy assessment of the methods.

[Norton et al. (2022)](https://doi.org/10.3390/rs14122896) combined LiDAR and hyperspectral data in a data fusion approach for classification of semi-arid woody cover species and achieved accuracies ranging from 86% to 98% for five woody species

We'll see how far we can get without hyperspectral data since this data is less common

## Textural covariates

[Rodman et al. (2019)](https://doi.org/10.1002/ecs2.2594) described a method using a series of image processing steps to add supplemental information describing the texture and context surrounding each image pixel in their analysis of forest cover change based only on panchromatic imagery (i.e. "black and white"). These textural covariates may be helpful for identifying slash piles, especially if we only have RGB data.

first, we'll make a single panchromatic band (the mean of red, green, and blue values) from our RGB imagery

```{r}
rgb_to_pan_fn <- function(r, g, b) {
  pan <- (r + g + b)/3
  # replace all 0's with na
  pan <- terra::subst(pan, from = 0, to = NA)
  return(pan)
}
panchromatic_rast <- rgb_to_pan_fn(
  ortho_rast[[1]], ortho_rast[[2]], ortho_rast[[3]]
)
```

quick plot

```{r}
terra::plot(
  panchromatic_rast
  , col = scales::pal_grey(0, 1)(100)
  , axes=F, legend = F
)
```

nice, now we'll make the textural covariates described by [Rodman et al. (2019)](https://doi.org/10.1002/ecs2.2594)

>During these steps, we identified locally dark pixels (indicative of individual trees surrounded by bright grassland) and quantified local standard deviation in brightness using moving windows at multiple scales (sensu Coburn and Roberts 2004; Fig. 3; Appendix S2). We combined these two layers with the original grayscale imagery to create three-band composite imagery (Fig. 3d). The combination of brightness, local minima, and standard deviation emphasizes differences in spectral reflectance that facilitate the separation of different forest structures (e.g., individual trees and dense stands) from non-forested areas. (p. 7)

and from Appendix S2

>Our first metric identifies pixels that are darker than their surroundings (hereafter "local minima"). To identify local minima, we calculated the mean and standard deviation of pixel brightness at each scale (3-15 pixel windows). A pixel was considered to be darker than its surroundings if its brightness value was lower than the window mean minus two standard deviations. These calculations were made for each window size for each pixel, then summed across all window sizes (thus, local minima values range 0-7, where 7 indicates that a pixel is darker than its surroundings across all moving window sizes; Fig. 3b) 

>For the second metric, we calculated the sum of standard deviation values across all window sizes (Fig. 3c). Standard deviation of pixel brightness is a simple version of local texture that can improve forest classification in high-resolution imagery. While standard deviation is more computationally efficient than many other texture metrics (e.g., entropy based on greylevel co-occurrence matrices), it may still yield comparable performance in classification (Coburn and Roberts 2004). Standard deviation appeared helpful in separating dark forest stands (low brightness, moderate variance) from more homogenous areas (low brightness, low variance; i.e., water bodies, shadows). 

>We combined these three layers (brightness, local minima, and standard deviation) to create a three-band composite image (Fig. 3d)


```{r, include=FALSE, eval=TRUE}
round_to_nearest_odd <- function(x) {
  rounded_int <- round(x)

  # Step 2: Check if the rounded integer is already odd.
  is_odd <- (rounded_int %% 2 != 0)

  # Step 3: For numbers that rounded to an even integer, find the nearest odd.
  odd_down <- rounded_int - 1
  odd_up <- rounded_int + 1

  # Calculate the absolute distances from the original number 'x'
  dist_down <- abs(x - odd_down)
  dist_up <- abs(x - odd_up)

  # Step 4: Use ifelse for vectorized conditional logic.
  result <- ifelse(is_odd,
                   rounded_int, # If the initially rounded integer is odd, use it.
                   ifelse(dist_down < dist_up,
                          odd_down, # If 'odd_down' is strictly closer.
                          odd_up    # If 'odd_up' is closer or equidistant.
                   )
  )

  return(result)
}
# round_to_nearest_odd(c(3.5, 4.2, 6.1, 7))
```

```{r}
rast_rodmanetal2019_fn <- function(
    rast, windows_m = c(3,5,7,9,11,13,15), scale_to_1m = T
    , rast_nm = "panchromatic"
) {
  rast <- rast[[1]]
  
  # scale the windows based on resolution
  if(scale_to_1m){
    windows_m <- round_to_nearest_odd(
      (1/terra::res(rast)[1])*windows_m
    )
  }else{
    windows_m <- round_to_nearest_odd(windows_m)
  }
  
  # means
  r_means <- windows_m %>% 
    purrr::map(\(w)
      terra::focal(x = rast, w = w, fun = "mean", na.rm = T)
    )
  # r_means[[7]] %>% terra::plot(col = scales::pal_grey(0, 1)(100))
  
  # sds
  r_sds <- windows_m %>% 
    purrr::map(\(w)
      terra::focal(x = rast, w = w, fun = "sd", na.rm = T)
    )
  # r_sds[[5]] %>% terra::plot(col = scales::pal_grey(0, 1)(100))
  
  # local min
  local_min_fn <- function(rast, mean, sd) {
    as.numeric(rast < ( mean-(sd*2) ))
  }
  
  r_local_min <- 1:length(r_means) %>% 
    purrr::map(\(i)
      local_min_fn(rast = rast, mean = r_means[[i]], sd = r_sds[[i]])
    )
  # r_local_min[[5]] %>% terra::plot()
  
  # aggregate local min
  r_local_min_agg <- terra::rast(r_local_min) %>% 
    cumsum()
  
  # get the last of the local mins
  local_minima <- r_local_min_agg[[terra::nlyr(r_local_min_agg)]]
  
  # aggregate local sd
  r_sds_agg <- terra::rast(r_sds) %>% 
    cumsum()
  
  # get the last of the local mins
  local_sd <- r_sds_agg[[terra::nlyr(r_sds_agg)]]
  
  # composite
  composite <- c(rast, local_sd, local_minima)
  names(composite) <- c(rast_nm, "local_sd", "local_minima")
  
  # names
  names(local_minima) <- c("local_minima")
  names(local_sd) <- c("local_sd")
  
  return(list(
    local_minima = local_minima
    , local_sd = local_sd
    , composite = composite
  ))
}
```

### Raster Texture: Panchromatic

implement our `rast_rodmanetal2019_fn` function
```{r}
rodmanetal2019_panchromatic_rast <- rast_rodmanetal2019_fn(
  rast = panchromatic_rast
  , rast_nm = "panchromatic"
  , scale_to_1m = F
)
```

plot of "local minima" pixels that are darker than their surroundings

```{r}
# plot
terra::plot(
  rodmanetal2019_panchromatic_rast$local_minima
  , col = scales::pal_grey(0, 1)(8)
  , axes=F, legend = F
)
```

plot of standard deviation pixel brightness which is a simple version of local texture that can improve forest classification in high-resolution imagery

```{r}
terra::plot(
  rodmanetal2019_panchromatic_rast$local_sd
  , col = scales::pal_grey(0, 1)(100)
  , axes=F, legend = F
)
```

plot of these combined panchromatic, local minima, and standard deviation images were merged to create a three-band composite which can be segmented to identify areas of relatively homogeneous brightness, contrast, and variance

```{r}
terra::plotRGB(
  rodmanetal2019_panchromatic_rast$composite
  # , scale = c(255,7,520)
  , stretch = "hist", colNA = "transparent"
)
# add polys
terra::plot(
  slash_piles_polys %>% sf::st_transform(terra::crs(grvi_rast)) %>% terra::vect()
  , add = T, border = "white", col = NA
)
```

### Raster Texture: GRVI

implement our `rast_rodmanetal2019_fn` function
```{r}
rodmanetal2019_grvi_rast <- rast_rodmanetal2019_fn(
  rast = grvi_rast
  , rast_nm = "grvi"
  , scale_to_1m = F
)
```

plot of these combined GRVI

```{r}
terra::plotRGB(
  rodmanetal2019_grvi_rast$composite
  # , scale = c(255,7,520)
  , stretch = "hist", colNA = "transparent"
)
# add polys
terra::plot(
  slash_piles_polys %>% sf::st_transform(terra::crs(grvi_rast)) %>% terra::vect()
  , add = T, border = "white", col = NA
)
```

### Plot textural rasters

let's look at these textural rasters for some of the piles

```{r}
p_fn_temp <- function(
    rn
    , df = slash_piles_polys
    , composite_rast = rodmanetal2019_grvi_rast$composite
    , crs = terra::crs(ortho_rast)
    , my_title = ""
) {
  # scale the buffer based on the largest
    d_temp <- df %>%
      dplyr::arrange(tolower(comment), desc(diameter)) %>% 
      dplyr::slice(rn) %>% 
      sf::st_transform(crs)
    
    # plt classifier
    # convert to stars
    comp_st <- composite_rast %>%  
      terra::subset(subset = c(1,2,3)) %>%
      terra::crop(
        d_temp %>% sf::st_buffer(20) %>% terra::vect()
      ) %>% 
      # terra::aggregate(fact = 2, fun = "mean", na.rm = T) %>% 
      stars::st_as_stars()
    
    # convert to rgb
    comp_rgb <- stars::st_rgb(
      comp_st[,,,1:3]
      , dimension = 3
      , use_alpha = FALSE
      # , stretch = "histogram"
      , probs = c(0.002, 0.998)
      , stretch = "percent"
    )
    # ggplot
    comp_temp <- ggplot2::ggplot() +
      stars::geom_stars(data = comp_rgb[]) +
      ggplot2::scale_fill_identity(na.value = "transparent") + # !!! don't take this out or RGB plot will kill your computer
      ggplot2::geom_sf(data = d_temp, fill = NA, color = "white", lwd = 0.3) +
      ggplot2::scale_x_continuous(expand = c(0, 0)) +
      ggplot2::scale_y_continuous(expand = c(0, 0)) +
      ggplot2::labs(
        x = ""
        , y = ""
        , subtitle = my_title
      ) +
      ggplot2::theme_void() +
      ggplot2::theme(
        legend.position = "none" # c(0.5,1)
        , legend.direction = "horizontal"
        , legend.margin = margin(0,0,0,0)
        , legend.text = element_text(size = 8)
        , legend.title = element_text(size = 8)
        , legend.key = element_rect(fill = "white")
        # , plot.title = ggtext::element_markdown(size = 10, hjust = 0.5)
        , plot.title = element_text(size = 8, hjust = 0.5, face = "bold")
        , plot.subtitle = element_text(size = 6, hjust = 0.5, face = "italic")
      )
    
  plt_temp <- comp_temp
  return(list("plt"=plt_temp,"d"=d_temp))
}
# combine 3
plt_combine_temp <- function(
    rn
    , df = slash_piles_polys %>% dplyr::filter(!is.na(comment))
    , composite_rast1 = rodmanetal2019_panchromatic_rast$composite
    , title1 = "panchromatic composite"
    , composite_rast2 = rodmanetal2019_grvi_rast$composite
    , title2 = "GRVI composite"
    , crs = terra::crs(ortho_rast)
) {
  # composite 1
  ans1 <- p_fn_temp(
    rn = rn
    , df = df
    , composite_rast = composite_rast1
    , my_title = title1
    , crs = crs
  )
  # composite 2
  ans2 <- p_fn_temp(
    rn = rn
    , df = df
    , composite_rast = composite_rast2
    , my_title = title2
    , crs = crs
  )
  # plt rgb
    rgb_temp <-
      ortho_plt_fn(ans1$d) + 
      ggplot2::geom_sf(data = ans1$d, fill = NA, color = "white", lwd = 0.3)
  # combine
    r <- ans1$plt + ans2$plt + rgb_temp
    return(r)
}

# add pile locations
plt_list_grvi_comp <- sample(1:nrow(slash_piles_polys %>% dplyr::filter(!is.na(comment))),10) %>% 
  purrr::map(plt_combine_temp)
# plt_list_grvi_comp[[2]]
```

combine the plots for a few piles

```{r, height = 10.5, width = 5}
patchwork::wrap_plots(
  plt_list_grvi_comp
  , ncol = 1
)
ggplot2::ggsave("../data/pile_tiles_texture.jpeg", height = 10.5, width = 5)
```

### Features for classification

[Goncalves et al. (2019)](https://doi.org/10.1016/j.jag.2018.11.011) describe the general workflow of the OBIA approach for ecological applications based on Genetic Algorithms (GA) to optimize image segmentation parameters:

1. Run the image segmentation: involves the partitioning of an image into a set of jointly exhaustive and mutually disjoint regions (a.k.a. segments, composed by multiple image pixels) that are internally more homogeneous and similar compared to adjacent ones;
    + segmentation algorithms include: mean-shift; region growing; SNIC (Simple Non-iterative Clustering) and SLIC (Simple Linear Iterative Clustering)
2. Load training data into the segmented image;
3. Calculate segment statistics (e.g., mean, standard-deviation) for classification features;
4. Merge training and segment statistics data (steps 2-3);
5. Do train/test data partitions (e.g., 5-fold cross-validation);
6. For each train/test set:
    6.1. Train the supervised classifier;
    6.2. Evaluate results for the test set;
7. Return the average evaluation score across all train/test rounds (fitness value);

where: 

* Segmentation features (step 1): a multi-layer raster dataset with features used only for the segmentation stage (e.g., spectral bands)
* Classification features (step 3): a multi-layer raster dataset with features used for classification (e.g., spectral data, band ratios, spectral vegetation indices, texture features, topographic data).

let's combine our rasters into a raster stack and generate a data frame with individual cells as rows and covariates as columns

```{r}
#rename composites
names(rodmanetal2019_panchromatic_rast$composite) <- c(
  "panchromatic", "panchromatic_local_sd", "panchromatic_local_minima"
)
names(rodmanetal2019_grvi_rast$composite) <- c(
  "grvi", "grvi_local_sd", "grvi_local_minima"
)
# raster stack
pred_rast <- c(
  terra::subset(ortho_rast, c("red","green","blue"))
  , rodmanetal2019_panchromatic_rast$composite
  , rodmanetal2019_grvi_rast$composite
)
```

check out correlations between covariates

```{r}
# investigate correlation among covariates
pred_rast %>% 
  terra::pairs(
    maxcells = min(7777, terra::ncell(pred_rast)*.01)
  )
```

```{r, include = F, eval = F}
# corrplot::corrplot.mixed(
corrplot::corrplot(
  pred_df %>% 
    dplyr::select(-c(cell,x,y)) %>% 
    dplyr::filter(dplyr::if_all(dplyr::everything(), ~ .x != -99)) %>% 
    stats::cor()
  , method = "ellipse"
  , order = "original" # "alphabet"
  , type = "lower"
  , diag = FALSE
)

pred_df %>% 
  # dplyr::filter(dplyr::if_all(dplyr::everything(), ~ .x != -99)) %>% 
  nrow()

pred_df %>% 
  dplyr::slice_sample(prop = 0.01) %>% 
  dplyr::filter(dplyr::if_all(dplyr::everything(), ~ .x != -99)) %>% 
  ggplot(aes(x = green, y = grvi)) +
  geom_point() + 
  geom_smooth(se=F, method = "lm")
```

the panchromatic band itself doesn't add much new information, let's remove it and generate a data frame for training our clustering algorithms

```{r}
# remove correlated layers
pred_rast <- terra::subset(
  pred_rast
  , subset = c("panchromatic")
  , negate = T
)
# combine rasters as data frame
pred_df <- terra::as.data.frame(
    pred_rast
    , xy = T
    , cell = T
  ) %>% 
  dplyr::mutate(
    dplyr::across(
      .cols = c(red,green,blue)
      , .fns = ~ ifelse(.x==0,NA,.x)
    )
  )
# what is this data?
pred_df %>% dplyr::slice_sample(n=100) %>% dplyr::glimpse()
```

## Pixel-based: unsupervised classification

Unsupervised image classification could be used to identify slash piles by grouping pixels with similar spectral characteristics. We'll call any method that implements classification at the pixel level like the unsupervised classification approach tested here "pixel-based image analysis" versus to OBIA. Unsupervised image classification uses algorithms like K-means to find natural clusters within spectral data and can use other inputs like elevation or spectral indices. This method operates without requiring pre-defined training data, making it useful when field data is limited. The output assigns each pixel to a class, which the user then interprets to potentially identify slash piles based on their distinct spectral signatures and elevated height. While effective for initial classification and pattern recognition, unsupervised methods can produce numerous classes that require manual interpretation and may be less accurate than supervised approaches without ground truth data.

let's test out kmeans clustering and see what we get

```{r}
set.seed(111)
# Create 6 clusters, allow 500 iterations, start with 5 random sets using "Lloyd"
kmncluster <- stats::kmeans(
  pred_df %>% 
    dplyr::select(-c(cell,x,y)) %>% 
    # we can't have NA values in kmeans
    dplyr::mutate(
      dplyr::across(
        .cols = dplyr::everything()
        , .fns = ~ as.numeric(ifelse(is.na(.x),-99,.x))
      )
    )
  , centers=6
  , iter.max = 500
  , nstart = 5
  , algorithm="Lloyd"
)
# kmeans returns an object of class kmeans
str(kmncluster)
```

put the kmeans clusters into a raster

```{r}
# make a blank raster
kmeans_cluster_rast <- terra::rast(ortho_rast, nlyr=1)
# fill the raster
kmeans_cluster_rast[pred_df$cell] <- kmncluster$cluster
# what?
kmeans_cluster_rast
```

there should be at most 6 clusters

```{r}
terra::freq(kmeans_cluster_rast)
```

plot the kmeans unsupervised classfication result

```{r}
terra::plot(
  kmeans_cluster_rast
  , col = viridis::turbo(n=6)
  , axes=F, legend = F
)
# add polys
terra::plot(
  slash_piles_polys %>% sf::st_transform(terra::crs(grvi_rast)) %>% terra::vect()
  , add = T, border = "white", col = NA
)

```

neat, let's see if the piles are classified by making a custom function to plot the classified raster and RGB side-by-side

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

```{r}
p_fn_temp <- function(
    rn
    , df = slash_piles_polys %>% dplyr::filter(!is.na(comment))
    , cluster_rast = kmeans_cluster_rast
    , crs = terra::crs(ortho_rast)
) {
  # scale the buffer based on the largest
    d_temp <- df %>%
      dplyr::arrange(tolower(comment), desc(diameter)) %>% 
      dplyr::slice(rn) %>% 
      sf::st_transform(crs)
    # plt
    rgb_temp <-
      ortho_plt_fn(d_temp) + 
      ggplot2::geom_sf(data = d_temp, fill = NA, color = "white", lwd = 1)
    
    # plt classifier
    class_temp <- ggplot2::ggplot() +
      ggplot2::geom_tile(
        data = cluster_rast %>%  
          terra::crop(
            d_temp %>% sf::st_buffer(20) %>% terra::vect()
          ) %>% 
          terra::as.data.frame(xy=T) %>% 
          dplyr::rename(f=3)
        , ggplot2::aes(x=x,y=y,fill=as.factor(f))
      ) +
      ggplot2::geom_sf(data = d_temp, fill = NA, color = "white", lwd = 1) +
      ggplot2::scale_fill_manual(
        values = c(
          "1" = viridis::turbo(n=6)[1]
          , "2" = viridis::turbo(n=6)[2]
          , "3" = viridis::turbo(n=6)[3]
          , "4" = viridis::turbo(n=6)[4]
          , "5" = viridis::turbo(n=6)[5]
          , "6" = viridis::turbo(n=6)[6]
          # , "7" = viridis::turbo(n=10)[7]
          # , "8" = viridis::turbo(n=10)[8]
          # , "9" = viridis::turbo(n=10)[9]
          # , "10" = viridis::turbo(n=10)[10]
        )
        , na.value = "transparent"
      ) +
      ggplot2::scale_x_continuous(expand = c(0, 0)) +
      ggplot2::scale_y_continuous(expand = c(0, 0)) +
      ggplot2::labs(
        x = ""
        , y = ""
        , fill = ""
      ) +
      ggplot2::theme_void() + 
      ggplot2::theme(
        legend.position = "left" # c(0.5,1)
        , legend.margin = margin(0,0,0,0)
        , legend.text = element_text(size = 6)
        , legend.title = element_text(size = 6)
        , legend.key = element_rect(fill = "white")
        # , plot.title = ggtext::element_markdown(size = 10, hjust = 0.5)
        , plot.title = element_text(size = 8, hjust = 0.5, face = "bold")
        , plot.subtitle = element_text(size = 6, hjust = 0.5, face = "italic")
      )
  plt_temp <- class_temp+rgb_temp
  return(plt_temp)
}
# add pile locations
plt_list_kmeans <- sample(1:nrow(slash_piles_polys %>% dplyr::filter(!is.na(comment))),10) %>% 
  purrr::map(p_fn_temp)
```

combine the plots for a few piles

```{r, height = 10.5, width = 8}
patchwork::wrap_plots(
  plt_list_kmeans
  , ncol = 2
  , guides = "collect"
)
ggplot2::ggsave("../data/pile_tiles_kmeans6.jpeg", height = 10.5, width = 8)
```

there is no discernible pattern between the kmeans classes and the slash piles because the lighting and conditions around the piles are not consistent. this result also highlights the challenges we'll likely face with pixel-based classification methods

## Pixel-based: supervised classification

we'll start with a pixel-based classification using the image-annotated pile polygons to train a classification model

first, let's set aside some piles for use in validation

```{r}
set.seed(333)
# 20% will be used for validation
slash_piles_polys <- slash_piles_polys %>% 
  dplyr::left_join(
    slash_piles_polys %>% 
      sf::st_drop_geometry() %>% 
      dplyr::filter(!is.na(comment)) %>% 
      dplyr::distinct(pile_id) %>%
      dplyr::slice_sample(prop = 0.2) %>% 
      dplyr::mutate(is_training = F)
    , by = "pile_id"
  ) %>% 
  dplyr::mutate(is_training = dplyr::coalesce(is_training, T))
slash_piles_points <- slash_piles_points %>% 
  dplyr::left_join(
    slash_piles_polys %>% 
      sf::st_drop_geometry() %>% 
      dplyr::distinct(objectid, is_training)
    , by = "objectid"
  )
# slash_piles_points %>% sf::st_drop_geometry() %>% dplyr::count(is_training)
```

first, we'll generate a set of sample points to extract raster values from to build our training data with covariates

```{r}
training_points <- dplyr::bind_rows(
# sample only from non-pile parts of the raster
  terra::mask(
    grvi_rast
    , slash_piles_polys %>% # use ALL piles (image-annotated and field-collected)
        sf::st_transform(terra::crs(grvi_rast)) %>% 
        terra::vect()
    , inverse = T
  ) %>%
  terra::spatSample(size = 5000, na.rm = T, xy = T) %>% 
  sf::st_as_sf(coords = c("x", "y"), crs = terra::crs(grvi_rast)) %>% 
  dplyr::mutate(
    is_pile = 0
  ) %>% 
  dplyr::select(is_pile)
# sample points within piles
  , slash_piles_polys %>% 
    dplyr::filter(
      is_training
    ) %>% 
    sf::st_transform(terra::crs(grvi_rast)) %>% 
    terra::vect() %>% 
    terra::spatSample(size = 1000) %>% 
    sf::st_as_sf() %>% 
    dplyr::mutate(
      is_pile = 1
    ) %>% 
    dplyr::select(is_pile)
) %>% 
dplyr::mutate(
  id = dplyr::row_number()
  , is_pile = as.factor(is_pile)
  , x = sf::st_coordinates(.)[,1]
  , y = sf::st_coordinates(.)[,2]
)
####### validation points since we'll validate at the pixel level
validation_points <- dplyr::bind_rows(
# sample only from non-pile parts of the raster
  terra::mask(
    grvi_rast
    , slash_piles_polys %>% # use ALL piles (image-annotated and field-collected)
        sf::st_transform(terra::crs(grvi_rast)) %>% 
        terra::vect()
    , inverse = T
  ) %>%
  # mask the training points as well
  terra::mask(
    training_points %>% terra::vect()
    , inverse = T
  ) %>% 
  terra::spatSample(size = 5000, na.rm = T, xy = T) %>% 
  sf::st_as_sf(coords = c("x", "y"), crs = terra::crs(grvi_rast)) %>% 
  dplyr::mutate(
    is_pile = 0
  ) %>% 
  dplyr::select(is_pile)
# sample points within piles
  , slash_piles_polys %>% 
    dplyr::filter(
      !is_training
    ) %>% 
    sf::st_transform(terra::crs(grvi_rast)) %>% 
    terra::vect() %>% 
    terra::spatSample(size = 1000) %>% 
    sf::st_as_sf() %>% 
    dplyr::mutate(
      is_pile = 1
    ) %>% 
    dplyr::select(is_pile)
) %>% 
dplyr::mutate(
  id = dplyr::row_number()
  , is_pile = as.factor(is_pile)
  , x = sf::st_coordinates(.)[,1]
  , y = sf::st_coordinates(.)[,2]
)
```

plot our RGB with training piles (white), validation piles (black), pile sample points (red), and non-pile sample points (blue)

```{r}
# let's check this out
terra::plotRGB(ortho_rast, stretch = "lin", colNA = "transparent")
terra::plot(
  slash_piles_polys %>% 
    dplyr::filter(is_training) %>% 
    sf::st_transform(terra::crs(ortho_rast)) %>% terra::vect()
  , add = T, border = "white", col = NA, lwd = 1.2
)
terra::plot(
  slash_piles_polys %>% 
    dplyr::filter(!is_training) %>% 
    sf::st_transform(terra::crs(ortho_rast)) %>% terra::vect()
  , add = T, border = "black", col = NA, lwd = 1.2
)
terra::plot(training_points %>% dplyr::filter(is_pile==1) %>% terra::vect(), col = "red", cex = 0.5, add = T)
terra::plot(training_points %>% dplyr::filter(is_pile==0) %>% terra::vect(), col = "blue", cex = 0.5, add = T)
```

extract the raster cell values from each layer in our RGB-based raster data, these values will be the predictor variables

```{r}
#############################
#extract covariate data at point locations
#############################
training_dta <-
  terra::extract(
    pred_rast
    , training_points %>%  
      terra::vect()
    , ID = F
  ) %>%  
  dplyr::bind_cols(
    training_points %>%  
      dplyr::select(id,is_pile,x,y)
  ) %>%  
  ## make sure no data is missing
  dplyr::filter(dplyr::if_all(dplyr::everything(), ~ !is.na(.x)))
  # ## square continuous vars?
  # dplyr::mutate(dplyr::across(
  #   .cols = dplyr::where(is.numeric) & !c(id, is_pile, x, y) & !tidyselect::ends_with("local_minima")
  #   , .fns = list(sq)
  #   , .names = "{.col}_sq" 
  # ))
##### and validation data to perform accuracy assesment at the pixel level
validation_dta <-
  terra::extract(
    pred_rast
    , validation_points %>%  
      terra::vect()
    , ID = F
  ) %>%  
  dplyr::bind_cols(
    validation_points %>%  
      dplyr::select(id,is_pile,x,y)
  ) %>%  
  ## make sure no data is missing
  dplyr::filter(dplyr::if_all(dplyr::everything(), ~ !is.na(.x)))
  # ## square continuous vars?
  # dplyr::mutate(dplyr::across(
  #   .cols = dplyr::where(is.numeric) & !c(id, is_pile, x, y) & !tidyselect::ends_with("local_minima")
  #   , .fns = list(sq)
  #   , .names = "{.col}_sq" 
  # ))
```

compare training and validation data to ensure that they are similar (i.e. we drew a big enough sample)

```{r}
dplyr::bind_rows(
    training_dta %>% sf::st_drop_geometry() %>% dplyr::mutate(dta_set = "Training")
    , validation_dta %>% sf::st_drop_geometry() %>% dplyr::mutate(dta_set = "Validation")
  ) %>% 
  dplyr::select(-c(geometry,dplyr::ends_with("_sq"))) %>% 
  dplyr::mutate_if(is.factor, as.numeric) %>% 
  tidyr::pivot_longer(
    cols = -c(id, is_pile, dta_set)
    , names_to = "covariate"
    , values_to = "value"
  ) %>% 
  dplyr::mutate(covariate = stringr::str_replace_all(covariate,"_", " ")) %>% 
ggplot(mapping = aes(x = value, color = dta_set, fill = dta_set, group = dta_set)) +
  geom_density(alpha=0.6) +
  facet_wrap(
    facets = vars(covariate), scales = "free"
    , labeller = label_wrap_gen(width = 25, multi_line = TRUE)
  ) + 
  scale_color_viridis_d(option = "cividis") +
  scale_fill_viridis_d(option = "cividis") +
  labs(
    title = "Distribution of Model Covariates"
    , subtitle = paste0(
      "by Training (n="
      , nrow(training_dta) %>% scales::comma(accuracy = 1)
      , ") and Validation (n="
      , nrow(validation_dta) %>% scales::comma(accuracy = 1)
      , ") data"
    )
  ) +
  theme_light() + 
  theme(
    legend.position = "top"
    , legend.title = element_blank()
    , plot.title = element_text(size = 10)
    , plot.subtitle = element_text(size = 9)
    , axis.title = element_text(size = 7)
    , axis.text = element_text(size = 6)
  ) + 
  guides(color = guide_legend(override.aes = list(size = 5)))
```

we'll use a random forest model to predict raster values

```{r}
# Tune on the full data if below threshold
tune_result_temp <- randomForest::tuneRF(
  y = training_dta$is_pile
  , x = training_dta %>% sf::st_drop_geometry() %>% dplyr::select(-c(id,is_pile,geometry))
  , ntreeTry = 500
  , stepFactor = 1
  , improve = 0.03
  , plot = F
  , trace = F
)
# Extract the optimal mtry value
optimal_mtry_temp <- tune_result_temp %>%
  dplyr::as_tibble() %>%
  dplyr::filter(OOBError==min(OOBError)) %>%
  dplyr::filter(dplyr::row_number() == 1) %>%
  dplyr::pull(mtry) %>%
  as.numeric()
# pass to random forest model
mod_rf <- randomForest::randomForest(
  y = training_dta$is_pile
  , x = training_dta %>% sf::st_drop_geometry() %>% dplyr::select(-c(id,is_pile,geometry))
  , mtry = optimal_mtry_temp
  , na.action = na.omit
  , ntree = 500
)
```

random forest variable importance plot

```{r}
#variable importance plot
randomForest::varImpPlot(mod_rf)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

predict raster values

```{r}
# x,y rasters
y_rast_temp <- terra::init(pred_rast[[1]], "y")
names(y_rast_temp) <- "y"
x_rast_temp <- terra::init(pred_rast[[1]], "x")
names(x_rast_temp) <- "x"
# predict
rf_pred_rast <- terra::predict(
  object = c(
      pred_rast
      , x_rast_temp, y_rast_temp
    )
  , model = mod_rf
  , na.rm = T
  # , type = "prob"
)
terra::coltab(rf_pred_rast) <- data.frame(value=1:2, col=c("gray", "khaki"))
levels(rf_pred_rast) <- data.frame(value=1:2, col=c("non-pile", "pile"))
```

plot pixel-level predictions

```{r}
#plot
terra::plot(rf_pred_rast)
terra::plot(
  slash_piles_polys %>% 
    dplyr::filter(is_training) %>% 
    sf::st_transform(terra::crs(ortho_rast)) %>% terra::vect()
  , add = T, border = "white", col = NA, lwd = 1.2
)
terra::plot(
  slash_piles_polys %>% 
    dplyr::filter(!is_training) %>% 
    sf::st_transform(terra::crs(ortho_rast)) %>% terra::vect()
  , add = T, border = "black", col = NA, lwd = 1.2
)
```

evaluate model performance based on the Area under the Receiver Operating Characteristic (ROC) Curve (AUC) criterion (an AUC value of 0.5 can be interpreted as the model performing no better than a random prediction). Higher AUC values indicate that the model is better at predicting "absence" classes as "absence" (true negative) and "presence" classes as "presence" (true positive).

```{r}
validation_dta <-
  validation_dta %>% 
  dplyr::bind_cols(
    terra::extract(
      rf_pred_rast
      , validation_points %>% terra::vect()
      , ID = F
    ) %>% 
    dplyr::rename(is_pile_predicted=1)
  )

PresenceAbsence::auc.roc.plot(
  validation_dta %>% 
    sf::st_drop_geometry() %>% 
    dplyr::select(id, is_pile, is_pile_predicted) %>% 
    dplyr::mutate(dplyr::across(tidyselect::starts_with("is_pile"), ~ as.numeric(.x)-1 ))
  , main = "Random Forest"
)
```

now lets check the confusion matrix of the pixel-based approach

```{r}
confusion_matrix_temp <-
  validation_dta %>% 
  dplyr::mutate(dplyr::across(tidyselect::starts_with("is_pile"), ~ as.numeric(.x)-1 )) %>% 
  dplyr::rename(
    presence = is_pile
    , estimate = is_pile_predicted
  ) %>% 
  dplyr::count(presence,estimate) %>% 
  dplyr::mutate(
    is_false = as.factor(ifelse(presence!=estimate,1,0))
    , presence_fact = factor(presence,levels = 0:1,labels = c("Observed Absent", "Observed Present"))
    , estimate_fact = factor(estimate,levels = 0:1,labels = c("Predicted Absent", "Predicted Present"))
    , pct = n/sum(n)
  )
# first function takes df with cols tp_n, fp_n, and fn_n to calculate rate
confusion_matrix_scores_fn <- function(df) {
  df %>% 
  dplyr::mutate(
    omission_rate = dplyr::case_when(
      (tp_n+fn_n) == 0  ~ as.numeric(NA)
      , T ~ fn_n/(tp_n+fn_n)
    ) # False Negative Rate or Miss Rate
    , commission_rate = dplyr::case_when(
      (tp_n+fp_n) == 0  ~ as.numeric(NA)
      , T ~ fp_n/(tp_n+fp_n)
    ) # False Positive Rate
    , precision = dplyr::case_when(
      (tp_n+fp_n) == 0  ~ as.numeric(NA)
      , T ~ tp_n/(tp_n+fp_n)
    )
    , recall = dplyr::case_when(
      (tp_n+fn_n) == 0  ~ as.numeric(NA)
      , T ~ tp_n/(tp_n+fn_n)
    )
    , f_score = dplyr::case_when(
      is.na(precision) | is.na(recall) ~ as.numeric(NA)
      , (precision+recall) == 0 ~ 0
      , T ~ 2 * ( (precision*recall)/(precision+recall) )
    )    
  ) 
}
confusion_matrix_scores_temp <- confusion_matrix_scores_fn(
  dplyr::tibble(
    tp_n = confusion_matrix_temp %>% dplyr::filter(presence==1 & estimate == 1) %>% dplyr::pull(n)
    , fn_n = confusion_matrix_temp %>% dplyr::filter(presence==1 & estimate == 0) %>% dplyr::pull(n)
    , fp_n = confusion_matrix_temp %>% dplyr::filter(presence==0 & estimate == 1) %>% dplyr::pull(n)
  )
)
# plot
ggplot(data =  confusion_matrix_temp, mapping = aes(y = estimate_fact, x = presence_fact)) +
  geom_tile(aes(fill = is_false), color = "white",alpha=0.8) +
  geom_text(aes(label = scales::comma(n,accuracy=1)), vjust = 1,size = 8) +
  geom_text(aes(label = scales::percent(pct,accuracy=0.1)), vjust = 3.5, size=5) +
  scale_fill_manual(values= c("turquoise","tomato2")) +
  scale_x_discrete(position = "top") +
  labs(
    y = "Predicted"
    , x = "Observed"
    , subtitle = paste0(
      "True positive rate (recall) = "
        , confusion_matrix_scores_temp$recall %>% 
          scales::percent(accuracy = 0.1)
      , "\nPrecision (PPV) = "
        , confusion_matrix_scores_temp$precision %>% 
          scales::percent(accuracy = 0.1)
      , "\nF1-score = "
        , confusion_matrix_scores_temp$f_score %>% 
          scales::percent(accuracy = 0.1)
    )
  ) +
  theme_light() + 
  theme(
    legend.position = "none"
    , panel.grid = element_blank()
    , plot.title = element_text(size = 9)
    , plot.subtitle = element_text(size = 9)
  )
```

that's not very good as we could tell from the prediction map

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
remove(mod_rf, rf_pred_rast, rodmanetal2019_grvi_rast, rodmanetal2019_panchromatic_rast)
gc()
```

## Object-based image analysis (OBIA)

OBIA is a technique (or a set of techniques) used to analyze digital images that was developed relatively recently in comparison to ‘classic’ pixel-based image approaches ([Burnett and Blaschke 2003](https://doi.org/10.1016/S0304-3800(03)00139-X)). We will probably implement the OBIA approach using the `SegOptim` [package](https://github.com/joaofgoncalves/SegOptim) presented by [Goncalves et al. (2019)](https://doi.org/10.1016/j.jag.2018.11.011)

An alternative that does not require 3rd party software and works entirely within the R framework is the `OpenImageR` [package](https://mlampros.github.io/OpenImageR/index.html).

### Training data

Training data is described as "typically a single-layer SpatRaster (from terra package) containing samples for training a classifier. The labels, classes or categories should be coded as integers {0,1} for single-class problems or, {1,2,...,n} for multi-class problems." We just have a single-class problem of classification with two mutually exclusive levels: piles and non-piles. Other examples include: species presence/absence, burned/unburned, cut/uncut forest.

first, we'll make a raster covering the entire study area with cells classified based on the pile polygon locations

```{r}
class_rast <-
  ortho_rast[[1]] %>% 
  # terra::subst(from = 0, to = NA) %>% ## 0 values to NA which is specific to this data
  terra::clamp(lower=0,upper=0) %>% 
  terra::mask(
    slash_piles_polys %>% # use ALL piles (image-annotated and field-collected)
        sf::st_transform(terra::crs(ortho_rast)) %>% 
        terra::vect()
    , inverse = T
    , updatevalue = 1
  )
# terra::plot(class_rast)
```

now we'll make a grid of 0.25 ha to sample from to define our training versus validation areas. we'll implement simple random sampling from this 0.25 ha grid.

after we make the sample area, we'll sample from our classfied raster

```{r, include = F, eval=FALSE}
# example tutorial: https://segoptim.bitbucket.io/docs/mapping-an-invasive-species-with-object-based-image-analysis.html#input-image-data
# dl from : https://segoptim.bitbucket.io/docs/data/test_data.zip
# Change this!!! 
derp_temp <- "c:/Users/georg/Downloads/SegOptim_test_data"

# Path to raster data used for image segmentation
# In SEGM_FEAT directory
inputSegFeat.path <- file.path(derp_temp,"SEGM_FEAT/SegmFeat_WV2_b532.tif")
seg_temp <- terra::rast(inputSegFeat.path)

# Path to training raster data
# [0] Non-invaded areas [1] Acacia dealbata invaded areas
# In TRAIN_AREAS directory
trainData.path <- file.path(derp_temp,"TRAIN_AREAS/TrainAreas.tif")
train_temp <- terra::rast(trainData.path)
terra::plot(train_temp, colNA = "black")

terra::plot(seg_temp[[1]], col = viridis::cividis(n=100))
terra::plot(train_temp, alpha = 0.3, add = T)
terra::ext(seg_temp[[1]]) %>% terra::vect() %>% terra::plot(border = "blue", col = NA, lwd = 10)
terra::ext(train_temp) %>% terra::vect() %>% terra::plot(border = "red", col = NA, lwd = 6, add = T)
# Path to raster data used as classification features
# In CLASSIF_FEAT directory
classificationFeatures.path <- c(file.path(derp_temp,"CLASSIF_FEAT/ClassifFeat_WV2_NDIs.tif"),
                                 file.path(derp_temp,"CLASSIF_FEAT/ClassifFeat_WV2_SpectralBands.tif"))
terra::rast(classificationFeatures.path[1])[[1]] %>% terra::ext() %>% terra::vect() %>% terra::plot(border = "orange", col = NA, lwd = 4, add = T)
terra::rast(classificationFeatures.path[2])[[1]] %>% terra::ext() %>% terra::vect() %>% terra::plot(border = "gray", col = NA, lwd = 2, add = T)


seg_temp %>% terra::plotRGB(stretch = "hist", colNA = "transparent")
terra::rast(classificationFeatures.path[1]) %>% names()
terra::rast(classificationFeatures.path[2]) %>% names()
terra::rast(classificationFeatures.path[2]) %>%
  terra::plotRGB(r = 1, g = 2, b = 3, stretch = "hist", colNA = "transparent")

# all of this has taught me that: 
# 1) the Training data (0/1 raster), Segmentation features (raster stack), and Classification features (raster stack) all cover the same area
# 2) the Training data (0/1 raster) is NA in areas not sampled... since our entire area is "sampled" we'll try to NA the validation areas

### Goncalves et al. (2019) describe 
    # The segmentation feature space: encompassed less dimensions (in comparison to the classification feature space) to allow fast processing needed to run segmentation. For example, the Red (R), Green (G) and Blue (B) bands were employed
    # Classification features were many: (a) spectral bands (reflectance values); (b) alternative colour spaces (other than RGB but with better separability between components; e.g., XYZ, YUV); (c) multiple combinations of band ratios in the form: where and are any two different spectral bands; (d) vegetation and/or water indices (e.g., NDVI, EVI, NDWI); (e) multiple combinations of Normalized Difference Indices (NDI) in the form ; (f) texture operators such as Haralick (Haralick, 1979) (HL; e.g., energy, entropy, correlation, inertia), Local Statistics (LS; e.g., mean, variance)
```

```{r}
# generates the grid
sample_grid <- ortho_rast[[1]] %>% 
  terra::clamp(lower=0,upper=0) %>% 
  terra::aggregate(fact = sqrt(10000*0.25)/terra::res(ortho_rast[[1]])[1])
# determins training tiles from the grid
set.seed(66)
sample_grid <-
  terra::init(sample_grid, "cell") %>% 
  terra::subst(
    from = sample(1:terra::ncell(sample_grid), size = floor(terra::ncell(sample_grid)*0.7))
    , to = 1
    , others = NA
  )
# terra::plot(sample_grid) # emmental
# sample from the classified raster
class_rast_training <-
  class_rast %>% 
    terra::mask(terra::as.polygons(sample_grid))
```

check out the area we'll use for training with the area for validation dimmed

```{r}
terra::plot(class_rast_training)
terra::plot(class_rast, alpha = 0.2, add = T, axes=F, legend = F)
```

### `OpenImageR` package SLIC image segmentation{#slic}

the `OpenImageR` [package](https://github.com/mlampros/OpenImageR) makes the Simple Linear Iterative Clustering (SLIC) clustering algorithm ([Achanta et al. 2012](https://scholar.google.com/scholar?cluster=2726314889973535954&hl=en&as_sdt=0,6)) available in R. SLIC takes an image and segments it into "superpixels." These superpixels are compact, nearly uniform regions that adhere well to image boundaries. The key idea is that within a superpixel, pixels are likely to belong to the same object or land cover type. This is a crucial first step because subsequent classification will be performed on these superpixels, not individual pixels.

```{r}
# the e1071 package is a popular choice for SVM implementation in R, providing a straightforward interface for training and evaluation.
# remotes::install_github('mlampros/OpenImageR')
# remotes::install_github('mlampros/SuperpixelImageSegmentation')
library("OpenImageR")
library("SuperpixelImageSegmentation")
# list.files(system.file("tmp_images", package = "OpenImageR"))
# system.file("tmp_images", "slic_im.png", package = "OpenImageR") %>% 
#   terra::rast() %>% 
#   # terra::subset(1) %>% 
#   # terra::plot()
#   terra::plotRGB()

#### save files to disk as required by package
tempdir_temp <- tempdir()
fpath_train <- file.path(tempdir_temp, "train.tif")
fpath_seg <- file.path(tempdir_temp, "seg.tif")
fpath_class <- file.path(tempdir_temp, "class.tif")
# write
terra::writeRaster(class_rast_training, filename = fpath_train, overwrite = T)
terra::writeRaster(
  pred_rast %>% terra::subset(c("red","green","blue"))
  , filename = fpath_seg
  , overwrite = T
)
terra::writeRaster(
  pred_rast # includes indicies and textural as well as RGB
  , filename = fpath_class
  , overwrite = T
)
```

we'll implement the SLICO variation of SLIC because the user does not need to define the compactness parameter or try different values of it. SLICO adaptively chooses the compactness parameter for each superpixel differently. This generates regular shaped superpixels in both textured and non textured regions alike. The improvement comes with hardly any compromise on the computational efficiency - SLICO continues to be as fast as SLIC ([**How is SLICO different from SLIC?**](https://www.epfl.ch/labs/ivrl/research/slic-superpixels/#SLICO))

```{r}
#--------------
# "slico" method
#--------------
# readImage on our RGB raster for segmentation
im_temp <- OpenImageR::readImage(fpath_seg)
# str(im_temp)

## number of superpixels to use
## set this to extent/minimum expected pile area?
sp_temp <- floor(
    # raster area
    ( terra::cellSize(ortho_rast[[1]])[1]*terra::ncell(ortho_rast[[1]]) ) / 
    # expected pile area
      (
        slash_piles_polys %>% dplyr::mutate(area = sf::st_area(.) %>% as.numeric()) %>% 
        dplyr::pull(area) %>% 
        quantile(probs = 0.95)
      )
  ) %>% 
  max(200) # 200 = function default

# superpixels
superpixels_slico_ans <- OpenImageR::superpixels(
  input_image = im_temp
  # Either "slic" or "slico"
  , method = "slico" 
  # number of superpixels to use
  , superpixel = sp_temp
  # numeric value specifying the compactness parameter. 
    # The compactness parameter is needed only if method is "slic". The "slico" method adaptively chooses
    # the compactness parameter for each superpixel differently
  # , compactness = 20 
  # If TRUE then the resulted slic or slico data will be returned
  , return_slic_data = TRUE
  # If TRUE then the Lab data will be returned ( the Lab-colour format )
  , return_labels = TRUE
  # If not an empty string ("") then it should be a path to the output file with extension .bin 
  , write_slic = ""
  , verbose = TRUE
)
```

what did we get back?

```{r}
str(superpixels_slico_ans)
# superpixels_slico_ans$labels %>% length()
# superpixels_slico_ans$labels %>% unique() %>% length()
# superpixels_slico_ans$labels %>% dplyr::as_tibble() %>% ncol()
# superpixels_slico_ans$labels %>% dplyr::as_tibble() %>% nrow()
# superpixels_slico_ans$slic_data %>% str()
```

let's put the superpixels into a raster

```{r}
# use the ortho template so we get the projection and then fill with values
superpixels_slico_rast <- ortho_rast %>% terra::subset(1)
# fill raster with the superpixels
superpixels_slico_rast[] <- superpixels_slico_ans$labels %>% terra::rast()
superpixels_slico_rast <- terra::as.factor(superpixels_slico_rast)
# reset the colors
terra::coltab(superpixels_slico_rast) <- data.frame(
  value = terra::minmax(superpixels_slico_rast)[1]:terra::minmax(superpixels_slico_rast)[2]
  , col = c(
    viridis::turbo(terra::unique(superpixels_slico_rast) %>% nrow() %>% `*`(1/3) %>% ceiling(), alpha = 0.35)
    , viridis::viridis(terra::unique(superpixels_slico_rast) %>% nrow() %>% `*`(1/3) %>% ceiling(), alpha = 0.35)
    , viridis::cividis(terra::unique(superpixels_slico_rast) %>% nrow() %>% `*`(1/3) %>% ceiling(), alpha = 0.35)
  ) %>% 
    sample(size = (terra::unique(superpixels_slico_rast) %>% nrow()))
)
# # plot
# superpixels_slico_rast %>% terra::plot(legend = F, axes = F)
```

let's look at the superpixels overlaid on the RGB image with the slash pile locations (blue)

```{r}
terra::plotRGB(ortho_rast, stretch = "lin", colNA = "black")
terra::plot(superpixels_slico_rast, add = T, alpha = 0.2, border = "white")
# add polys
terra::plot(
  slash_piles_polys %>% sf::st_transform(terra::crs(ortho_rast)) %>% terra::vect()
  , add = T, border = "blue", col = NA
)
# OpenImageR::NormalizeObject(superpixels_slico_ans$slic_data) %>% 
#   grDevices::as.raster() %>% 
#   graphics::plot()
```

that's neat, the superpixels are moderately aligned with the large machine piles but are generally too large for the smaller hand piles

the steps in the OBIA approach workflow described by [Goncalves et al. (2019)](https://doi.org/10.1016/j.jag.2018.11.011) would be:

1. Image Segmentation (e.g. using SLIC) as demonstrated above
2. Feature Extraction for Superpixels: instead of using individual pixel values, extract features that describe each superpixel (e.g. spectral bands, band ratios, textural features)
3. Use labeled training data to assign superpixels to their respective classes and perform classifier training (i.e. using random forest, support vector machines, etc.)
4. Once the classifier is trained, you apply it to the features of all the unlabeled superpixels in the image

### `SuperpixelImageSegmentation` package SLICAP clustering

we'll test out the `SuperpixelImageSegmentation` [package](https://github.com/mlampros/SuperpixelImageSegmentation) which was developed specifically to implement the SLIC Superpixels and Affinity Propagation (AP) Clustering method of image segmentation ([Zhou 2015](https://www.ijsr.net/archive/v4i4/SUB152869.pdf)). This method first uses the SLIC superpixel algorithm to form an over-segmentation of an image, constructs a similarity score based on the features of superpixels, and uses the AP algorithm to cluster superpixels. All of these steps are performed in a semi-automated manner as the user does not need to specify the number of clusters as required by other methods (e.g. Kmeans). 

```{r}
# readImage on our RGB raster for segmentation
im_temp <- OpenImageR::readImage(fpath_seg)
# initiate and perform segmentation of images based on superpixels, affinity propagation and kmeans clustering
init <- SuperpixelImageSegmentation::Image_Segmentation$new()
superpixels_slicap_ans <- init$spixel_segmentation(
  input_image = im_temp
  # a numeric value specifying the number of superpixels
  , superpixel = 600
  # Either "slic" or "slico"
  , method = "slico"
  # a character string specifying the kmeans method. If not empty ("") then it can be either "kmeans" or "mini_batch_kmeans"
  , kmeans_method = ""
  # If TRUE then the affinity propagation image data will be computed and returned
  , AP_data = TRUE
  # If TRUE then the median will be used rather than the mean value for the inner computations (see the details section for more information)
  , use_median = TRUE
  # a numeric value specifying the weight for the "L" channel of the image (the weights of the three channels based on the "CIELAB color space")
  , sim_wL = 3
  # a numeric value specifying the weight for the "A" channel of the image (the weights of the three channels based on the "CIELAB color space")
  , sim_wA = 10
  # a numeric value specifying the weight for the "B" channel of the image (the weights of the three channels based on the "CIELAB color space")
  , sim_wB = 10
  # numeric value specifying the colorradius: 
    ## adjusts the number of clusters, and if its value is low, the number of targets would increase, which leads to more detailed segmentation results
  , sim_color_radius = 10
  , verbose = TRUE
)
```

what did we get back?

```{r}
str(superpixels_slicap_ans)
```

let's see the output based on **Superpixels + AP**

```{r}
OpenImageR::imageShow(superpixels_slicap_ans$AP_image_data)
```

ok, let's increase the number of `superpixel` and decrease the `sim_color_radius` to get finer detail (but significantly increase processing time)

```{r}
# readImage on our RGB raster for segmentation
im_temp <- OpenImageR::readImage(fpath_seg)
# initiate and perform segmentation of images based on superpixels, affinity propagation and kmeans clustering
init <- SuperpixelImageSegmentation::Image_Segmentation$new()
superpixels_slicap_ans <- init$spixel_segmentation(
  input_image = im_temp
  # a numeric value specifying the number of superpixels
  , superpixel = sp_temp # 3333
  # Either "slic" or "slico"
  , method = "slico"
  # a character string specifying the kmeans method. If not empty ("") then it can be either "kmeans" or "mini_batch_kmeans"
  , kmeans_method = ""
  # If TRUE then the affinity propagation image data will be computed and returned
  , AP_data = TRUE
  # If TRUE then the median will be used rather than the mean value for the inner computations (see the details section for more information)
  , use_median = TRUE
  # a numeric value specifying the weight for the "L" channel of the image (the weights of the three channels based on the "CIELAB color space")
  , sim_wL = 3
  # a numeric value specifying the weight for the "A" channel of the image (the weights of the three channels based on the "CIELAB color space")
  , sim_wA = 10
  # a numeric value specifying the weight for the "B" channel of the image (the weights of the three channels based on the "CIELAB color space")
  , sim_wB = 10
  # numeric value specifying the colorradius: 
    ## adjusts the number of clusters, and if its value is low, the number of targets would increase, which leads to more detailed segmentation results
  , sim_color_radius = 2
  , verbose = TRUE
)
```

let's see the clustering output based on **Superpixels + AP** and we'll add the pile outlines in blue

```{r, include = F, eval = F}
OpenImageR::imageShow(superpixels_slicap_ans$AP_image_data)
superpixels_slicap_ans %>% str()
superpixels_slicap_ans$AP_image_data %>% str()
superpixels_slicap_ans$AP_image_data[,,1] %>% c() %>% unique() %>% length()
superpixels_slicap_ans$AP_image_data[,,2] %>% c() %>% unique() %>% length()
superpixels_slicap_ans$AP_image_data[,,3] %>% c() %>% unique() %>% length()
```

```{r}
# use the ortho template so we get the projection and then fill with values
rast_temp <- ortho_rast %>% terra::subset(1)
rast_temp[] <- superpixels_slicap_ans$AP_image_data[,,1] %>% terra::rast()
# the unique values in the layer represent the unique clusters
rast_temp <- terra::subst(
    rast_temp
    , from = terra::values(rast_temp) %>% unique() %>% c()
    , to = 1:(terra::values(rast_temp) %>% unique() %>% length())
  ) %>% 
  terra::as.factor()
# rast_temp %>% terra::crs()
# rast_temp %>% terra::plot()
# reset the colors
terra::coltab(rast_temp) <- data.frame(
  value = terra::minmax(rast_temp)[1]:terra::minmax(rast_temp)[2]
  , col = viridis::turbo(terra::unique(rast_temp) %>% nrow(), alpha = 0.9)
)
# plot
terra::plot(rast_temp, colNA = "black", axes=F)
terra::plot(
  slash_piles_polys %>% 
    sf::st_transform(terra::crs(ortho_rast)) %>% terra::vect()
  , add = T, border = "blue", col = NA, lwd = 1.2, axes=F
)
```

it looks like the automatic clustering did not consistently classify the pile locations. let's check the distribution of the clusters within the pile locations

```{r}
tab_temp <- rast_temp %>% 
  terra::extract(
    slash_piles_polys %>% 
      sf::st_transform(terra::crs(ortho_rast)) %>% terra::vect()
    , ID = F
  ) %>% 
  dplyr::rename(cluster = 1) %>% 
  dplyr::count(cluster) %>% 
  dplyr::mutate(pct = n/sum(n)) %>% 
  dplyr::arrange(desc(n)) 
# table
tab_temp %>% 
  dplyr::select(-n) %>% 
  dplyr::mutate(pct = scales::percent(pct,accuracy=0.1)) %>% 
  kableExtra::kbl(
    col.names = c("cluster","pct of clusters<br>that overlap with a pile")
    , escape = F
  ) %>% 
  kableExtra::kable_styling()
```

we can see this if we look at only the clusters that overlap with at least 10% of the piles

```{r}
### 
terra::subst(
  rast_temp
  , from = tab_temp %>% dplyr::filter(pct>=0.1) %>% dplyr::pull(cluster) %>% as.numeric()
  , to = tab_temp %>% dplyr::filter(pct>=0.1) %>% dplyr::pull(cluster) %>% as.numeric()
  , others = NA
) %>% 
  terra::plot(axes = F)
terra::plot(
  slash_piles_polys %>% 
    sf::st_transform(terra::crs(ortho_rast)) %>% terra::vect()
  , add = T, border = "blue", col = NA, lwd = 1.2, axes=F
)
```

this testing indicated that clustering based on **Superpixels + AP** will likely not work for our problem to detect slash piles in imagery because the output is not granular enough

#### Kmeans to automatically determine clusters

one might take advantage of the automatic way the AP algorithm determines the number of clusters in order to perform vector quantization using kmeans (this will give qualitative better results at the cost of increasing computation time). the advantage of using Superpixels + AP + Kmeans (or Mini-Batch-Kmeans, which can decrease the computation time at the cost of a reduced image quality) lies in the fact that the user does not have to specify the number of clusters beforehand

to do this, we update the `kmeans_*` parameters

**this took 40 minutes with kmeans_method = "kmeans"**

```{r}
# initiate and perform segmentation of images based on superpixels, affinity propagation and kmeans clustering
# ?SuperpixelImageSegmentation::Image_Segmentation
init <- SuperpixelImageSegmentation::Image_Segmentation$new()
superpixels_slicap_ans <- init$spixel_segmentation(
  input_image = im_temp
  # a numeric value specifying the number of superpixels
  , superpixel = sp_temp # 3333
  # Either "slic" or "slico"
  , method = "slico"
  # a character string specifying the kmeans method. If not empty ("") then it can be either "kmeans" or "mini_batch_kmeans"
  , kmeans_method = "mini_batch_kmeans"
  # the method of initialization. One of, optimal_init, quantile_init, kmeans++ and random
  , kmeans_initializer = "kmeans++"
  # number of times the algorithm will be run with different centroid seeds
  , kmeans_num_init = 3
  # the maximum number of clustering iterations
  , kmeans_max_iters = 100
  # If TRUE then the affinity propagation image data will be computed and returned
  , AP_data = TRUE
  # If TRUE then the median will be used rather than the mean value for the inner computations (see the details section for more information)
  , use_median = TRUE
  # a numeric value specifying the weight for the "L" channel of the image (the weights of the three channels based on the "CIELAB color space")
  , sim_wL = 3
  # a numeric value specifying the weight for the "A" channel of the image (the weights of the three channels based on the "CIELAB color space")
  , sim_wA = 10
  # a numeric value specifying the weight for the "B" channel of the image (the weights of the three channels based on the "CIELAB color space")
  , sim_wB = 10
  # numeric value specifying the colorradius: 
    ## adjusts the number of clusters, and if its value is low, the number of targets would increase, which leads to more detailed segmentation results
  , sim_color_radius = 2
  , verbose = TRUE
)
```

what did we get back?

```{r}
superpixels_slicap_ans %>% str()
```

let's see the clustering output based on **Superpixels + AP + Kmeans** and we'll add the pile outlines in blue

```{r}
# # Superpixels + AP
# OpenImageR::imageShow(superpixels_slicap_ans$AP_image_data)
# # Superpixels + AP + Kmeans
# OpenImageR::imageShow(superpixels_slicap_ans$KMeans_image_data)
# use the ortho template so we get the projection and then fill with values
rast_temp <- ortho_rast %>% terra::subset(1)
rast_temp[] <- superpixels_slicap_ans$KMeans_image_data[,,1] %>% terra::rast()
# the unique values in the layer represent the unique clusters
rast_temp <- terra::subst(
    rast_temp
    , from = terra::values(rast_temp) %>% unique() %>% c()
    , to = 1:(terra::values(rast_temp) %>% unique() %>% length())
  ) %>% 
  terra::as.factor()
# rast_temp %>% terra::crs()
# rast_temp %>% terra::plot()
# reset the colors
terra::coltab(rast_temp) <- data.frame(
  value = terra::minmax(rast_temp)[1]:terra::minmax(rast_temp)[2]
  , col = viridis::turbo(terra::unique(rast_temp) %>% nrow(), alpha = 0.9)
)
# plot
terra::plot(rast_temp, colNA = "black", axes=F)
terra::plot(
  slash_piles_polys %>% 
    sf::st_transform(terra::crs(ortho_rast)) %>% terra::vect()
  , add = T, border = "blue", col = NA, lwd = 1.2, axes=F
)
```

let's check the distribution of the clusters within the pile locations

```{r}
tab_temp <- rast_temp %>% 
  terra::extract(
    slash_piles_polys %>% 
      sf::st_transform(terra::crs(ortho_rast)) %>% terra::vect()
    , ID = F
  ) %>% 
  dplyr::rename(cluster = 1) %>% 
  dplyr::count(cluster) %>% 
  dplyr::mutate(pct = n/sum(n)) %>% 
  dplyr::arrange(desc(n)) 
# table
tab_temp %>% 
  dplyr::select(-n) %>% 
  dplyr::mutate(pct = scales::percent(pct,accuracy=0.1)) %>% 
  kableExtra::kbl(
    col.names = c("cluster","pct of clusters<br>that overlap with a pile")
    , escape = F
  ) %>% 
  kableExtra::kable_styling()
```

did anything change with the **Superpixels + AP** output?

```{r}
# use the ortho template so we get the projection and then fill with values
rast_temp <- ortho_rast %>% terra::subset(1)
rast_temp[] <- superpixels_slicap_ans$AP_image_data[,,1] %>% terra::rast()
# the unique values in the layer represent the unique clusters
rast_temp <- terra::subst(
    rast_temp
    , from = terra::values(rast_temp) %>% unique() %>% c()
    , to = 1:(terra::values(rast_temp) %>% unique() %>% length())
  ) %>% 
  terra::as.factor()
# rast_temp %>% terra::crs()
# rast_temp %>% terra::plot()
# reset the colors
terra::coltab(rast_temp) <- data.frame(
  value = terra::minmax(rast_temp)[1]:terra::minmax(rast_temp)[2]
  , col = viridis::turbo(terra::unique(rast_temp) %>% nrow(), alpha = 0.9)
)
# plot
terra::plot(rast_temp, colNA = "black", axes=F)
terra::plot(
  slash_piles_polys %>% 
    sf::st_transform(terra::crs(ortho_rast)) %>% terra::vect()
  , add = T, border = "blue", col = NA, lwd = 1.2, axes=F
)
```

no.

### Feature Extraction for Superpixels

let's go back to classifying the superpixels we generated using [SLICO](#slic)

1. Image Segmentation (e.g. using SLIC) as [demonstrated above](#slic)
2. Feature Extraction for Superpixels: instead of using individual pixel values, extract features that describe each superpixel (e.g. spectral bands, band ratios, textural features)
3. Use labeled training data to assign superpixels to their respective classes and perform classifier training (i.e. using random forest, support vector machines, etc.)
4. Once the classifier is trained, you apply it to the features of all the unlabeled superpixels in the image

```{r}
# terra::plot(superpixels_slico_rast, legend = F)
# terra::values(superpixels_slico_rast) %>% unique() %>% length()
# terra::as.polygons(superpixels_slico_rast, aggregate = T, values = T) %>% terra::plot(legend = T)
# convert superpixels to polygons
superpixels_slico_sf <- terra::as.polygons(superpixels_slico_rast, aggregate = T, values = T) %>% 
  sf::st_as_sf() %>% 
  dplyr::rename(superpixel_id = 1)
# list of functions to aggregate pixel data to superpixel
fn_list <- list(
  mean = function(x){mean(x,na.rm = T)}
  , median = function(x){median(x,na.rm = T)}
  , sd = function(x){sd(x,na.rm = T)}
  , q90 = function(x){quantile(x, probs = 0.9, na.rm = T)}
  , q10 = function(x){quantile(x, probs = 0.1, na.rm = T)}
  , mode = function(x) {
      # Get the unique values in the vector
      unique_values <- unique(x)
      # Calculate the frequency of each unique value
      value_frequencies <- tabulate(match(x, unique_values))
      # Find the index of the maximum frequency
      max_frequency_index <- which.max(value_frequencies)[1]
      # Return the unique value corresponding to the maximum frequency
      return(unique_values[max_frequency_index])
    }
)
# feature extraction for superpixels
superpixels_features_df <-
  pred_rast %>% 
  terra::extract(
    y = superpixels_slico_sf %>% terra::vect()
    # , fun = "table"
    , ID = T
  ) %>% 
  dplyr::group_by(ID) %>% 
  dplyr::summarise(dplyr::across(
    .cols = dplyr::where(is.numeric)
    , .fns = fn_list
    , .names = "{.col}_{.fn}"
  )) %>% 
  dplyr::ungroup() %>% 
  # get rid of summary variables for local_minima values since it's on 0-8 scale
  dplyr::select(
    -c(
      tidyselect::contains("local_minima_mean")
      , tidyselect::contains("local_minima_median")
      , tidyselect::contains("local_minima_sd")
      , tidyselect::contains("local_minima_q10")
      , tidyselect::contains("local_minima_q90")
    )
  )

superpixels_features_df %>% 
   %>% 
  # dplyr::select(tidyselect::contains("local_minima")) %>% 
  # summary()

# superpixels_slico_sf %>% 
#   dplyr::bind_cols(superpixels_features_df) %>% 
#   ggplot() + geom_sf(aes(fill = grvi_median)) + 
#   harrypotter::scale_fill_hp(option = "slytherin", na.value = "black") +
#   theme_void() + theme(legend.position = "none")
# grvi_rast %>% terra::aggregate(fact = 4) %>%  
#   terra::plot(axes = F, legend = F, col = harrypotter::hp(n=200, option = "slytherin"))
```

